---
title: "Humpback distribution and sea ice"
author: "Ben Weinstein"
date: "February 21, 2017"
output: 
  html_document:
    keep_md: yes
    number_sections: yes
    theme: spacelab
    toc: yes
---

# Project Description 
  
  * Global changes in climate will lead to new environments
  
  * A key question is how sensitive animals are to changes in environmental conditions
  
  * When favorable conditions arise, how quickly can animals respond to the availability of new habitat?
  
  * Polar marine ecosystems are particularly vulnerable, given the pronounced effects of climate on local conditions, variable environments, and lack of information on species distributions
  
  * We evaluate the response of two cetacean species to changes in sea-ice cover.
  
##	Our aims

  * Sea ice trends, by month, over the last 5 years

  * Determine the threshold of sea-ice concentration and proximity on whale presence
  
  * Evaluate the phenology of sea-ice timing and whale movement
  
  * Highlight anomalous sea ice conditions  
  * % Change in available habitat based on variable ice conditions

```{r,warning=FALSE,message=FALSE,echo=FALSE}
library(knitr)
library(foreach)
library(doSNOW)
opts_chunk$set(echo=F,warning=F,message=F,fig.width = 11,fig.height = 7,cache=F)
library(boot)
library(raster)
library(tidyr)
library(ggplot2)
library(MASS)
library(ggmap)
library(dplyr)
library(chron)
library(gridExtra)
library(stringr)
library(R2jags)
library(maptools)
library(reshape2)
#New model to be run, flag turned off if just updating.
newModel<-T

##ggplot theme
mytheme<-theme(axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank(),axis.title.x=element_blank(),axis.title.y=element_blank(),panel.grid=element_blank())

#load("Data/IceData.RData")
daysec = 60*60*24
region='AntarcticPeninsula'
```

```{r}
#read data
mdat<-read.csv("Data/Humpback Whales Megaptera novaeangliae West Antarctic Peninsula-3343066988628153526.csv")
#standardize column names to match the simulation
#Create an animal tag.
mxy <- as(mdat, "data.frame")
mxy$Animal<-mxy$individual.local.identifier
mxy$x<-mxy$location.long
mxy$y<-mxy$location.lat

mxy$argos.lc<-mxy$argos.iq
mxy$argos.iq<-NULL

#grab set of animals
mxy<-mxy[mxy$Animal %in% c("131143","131142","123232","123236"),]

#get rid of z class 
mxy<-mxy[!mxy$argos.lc %in% c("B","Z"),]

#crop by extent
d<-SpatialPointsDataFrame(cbind(mxy$x,mxy$y),data=mxy,proj4string=CRS("+proj=longlat +datum=WGS84"))

cropoly<-readShapePoly("Data/CutPolygon.shp",proj4string=CRS("+proj=longlat +datum=WGS84"))

b<-d[!is.na(d %over% cropoly)[,2],]

mxy<-b@data

#set datestamp
mxy$timestamp<-as.POSIXct(mxy$timestamp,format="%Y-%m-%d %H:%M:%S.000",tz="GMT")

#month and year columns
mxy$Month<-months(mxy$timestamp)
mxy$Year<-years(mxy$timestamp)

#remove migration events, create user cut polygon
migration<-data.frame(Animal=c("121207","121210","123224","131130","123236","123232","112699","131127","121208","131136","131132","131133"),timestamp=c("2013-05-07 12:16:26","2013-04-30 02:01:51","2013-05-23 01:23:55","2016-04-27 00:53:21","2013-03-16 05:35:06","2013-04-25 05:09:21","2012-06-15 03:28:15","2016-07-15 00:02:46","2013-02-12 00:51:28","2016-06-30 00:18:11","2016-05-09 03:57:47","2016-07-05 13:26:44"))
migration$timestamp<-as.POSIXct(migration$timestamp,format="%Y-%m-%d %H:%M:%S")

#mxy %>% group_by(Animal) %>% summarize(max(timestamp,na.rm=T)) %>% filter(Animal %in% migration$Animal)

for(x in 1:nrow(migration)){
    toremove<-which(mxy$Animal %in% migration$Animal[x] & mxy$timestamp > migration$timestamp[x])
    mxy<-mxy[!(1:nrow(mxy) %in% toremove),]
}           

#Only austral sping and summer, not enough data for june and july
#mxy<-mxy[mxy$Month %in% month.name[1:6],]

#remove empty timestamps
mxy<-mxy[!is.na(mxy$timestamp),]

#remove duplicates
mxy<-mxy[!duplicated(data.frame(mxy$timestamp,mxy$Animal)),]
```

# Occupancy and ice concentration.

What is the probability of occupancy of a cell as a function of % ice cover.

Associate each argos location with ice cell.

```{r}

#get a sample raster
r<-raster("C:\\Users\\Ben\\Dropbox\\Whales\\AMSR2\\2016\\may\\AntarcticPeninsula\\asi-AMSR2-s3125-20160531.grd")

#get points in projection
longlat<-SpatialPoints(cbind(mxy$location.long,mxy$location.lat),proj4string=CRS("+proj=longlat +datum=WGS84"))
#long lat
polarstereo<-spTransform(x=longlat,CRSobj=crs(r))
mxy[,c("stereo.x","stereo.y")]<-coordinates(polarstereo)
```

```{r}
#unique days to pull raster
udays<-unique(strptime(mxy$timestamp,format="%Y-%m-%d",tz="GMT"))
mxy$iceday<-strptime(mxy$timestamp,format="%Y-%m-%d",tz="GMT")

#for each iceday, get the sea ice concentration
out<-list()
for (x in 1:length(udays)){
    #split into components
  mn<-format(udays[x],format="%m")
  
  #month abbreciation
  mnabb<-tolower(month.abb[as.numeric(mn)])
  
  day<-format(udays[x],format="%d")
  year<-format(udays[x],format="%Y")
  
  fl_final<-paste('asi-AMSR2-s3125-',year,mn,day,'.grd',sep="")
  loadfile<-paste("C:/Users/Ben/Dropbox/Whales/AMSR2",year,mnabb,region,fl_final,sep="/")
  
  #load raster
  tryCatch({
  r<-raster(loadfile)
  
  #extract pts
  upts<-mxy[mxy$iceday %in% udays[x],c("stereo.x","stereo.y")]
  out[[x]]<-data.frame(upts,ice=raster::extract(x=r,y=SpatialPoints(cbind(upts))))
  }, error=function(e){
    #print("No file")
    return(NA)
  })
}

ice_frame<-bind_rows(out)
mxy<-merge(mxy,ice_frame,by=c("stereo.x","stereo.y"))
```

```{r}
mxy$Month<-factor(mxy$Month,levels=month.name)
ggplot(mxy,aes(x=ice)) + geom_histogram(aes(fill=Month),position="dodge") + labs(x="Sea Ice Concentration") + theme_bw()

ggplot(mxy[mxy$ice>15,],aes(x=ice)) + geom_histogram(aes(fill=Month),position="dodge") + labs(x="Sea Ice Concentration") + theme_bw()

ggplot(mxy,aes(x=Month,y=ice)) + geom_boxplot() + labs(y="Sea Ice Concentration") + theme_bw()
```

```{r}
ggplot()+geom_point(data=mxy, aes(x=x, y=y,col=ice),size=.5)  + facet_wrap(~Month,nrow = 2)  + borders(fill="black") + coord_cartesian(ylim = c(-70,-62),xlim=c(-55,-73)) + theme_bw() + mytheme + scale_color_continuous(low="blue",high="red") 

ggplot()+geom_point(data=mxy[mxy$ice>15,], aes(x=x, y=y,col=ice),size=.5)  + facet_wrap(~Month,nrow = 2)  + borders() + coord_cartesian(ylim = c(-70,-62),xlim=c(-55,-73)) + theme_bw() + mytheme + scale_color_continuous(low="blue",high="red") 

```

##Time integration

Each observation is not independent. In the above analysis a whale popping up 4 times in an hour will recieve the same weight as 4 observations in one day. One approach is to take the average ice concentration among observations and multiply it be the time difference.

```{r}

difft<-function(df){
  df<-df %>% arrange(timestamp)
for(row in 2:nrow(df)){
    df[row,"LastTime"]<-difftime(df$timestamp[row],df$timestamp[row-1],units="hours")
     df[row,"MeanIce"]<-median(c(df$ice[row],df$ice[row-1]),na.rm=T)
  }
  return(df)
}
#time since last observation
mxy$iceday<-as.character(mxy$iceday)
mxy_time<-mxy %>% group_by(Animal) %>% do(difft(.))

#remove observations more than a two days, too far to travel
mxy_time<-mxy_time %>% filter(LastTime<48)


ggplot(mxy_time,aes(x=MeanIce))  + geom_density(fill="red") + labs(x="Sea Ice Concentration(%)") + theme_bw()

#weighted kernel
ggplot(mxy_time,aes(x=MeanIce))  + geom_density(fill="red") + geom_density(aes(weight=LastTime/sum(LastTime)),fill="black",alpha=0.3) + labs(x="Sea Ice Concentration(%)")
```

# Null models

## Random use

Null use map

For each day, sample randomly in the background points for null values of sea ice concentration.

```{r}
stereopts<-SpatialPoints(cbind(mxy$stereo.x,mxy$stereo.y),proj4string =crs(r))

background<-raster(r)
bg<-rasterize(stereopts,background)
bg<-bg>0

#for each day

#unique days to pull raster
udays<-unique(strptime(mxy$timestamp,format="%Y-%m-%d",tz="GMT"))
mxy$iceday<-strptime(mxy$timestamp,format="%Y-%m-%d",tz="GMT")

#for each iceday, get the sea ice concentration
out<-list()
for (x in 1:length(udays)){
    #split into components
  mn<-format(udays[x],format="%m")
  
  #month abbreciation
  mnabb<-tolower(month.abb[as.numeric(mn)])
  
  day<-format(udays[x],format="%d")
  year<-format(udays[x],format="%Y")
  
  fl_final<-paste('asi-AMSR2-s3125-',year,mn,day,'.grd',sep="")
  loadfile<-paste("C:/Users/Ben/Dropbox/Whales/AMSR2",year,mnabb,region,fl_final,sep="/")
  
  #load raster
  tryCatch({
  r<-raster(loadfile)
  
  #mask by bg points
  #r<-mask(r,bg)
  
  #how many points?
  bgpts<-sum(mxy$iceday %in% udays[x])
  out[[x]]<-sampleRandom(r,xy=T,size=bgpts)

  }, error=function(e){
    #print("No file")
    return(NA)
  })
}

nullframe<-bind_rows(lapply(out,function(x) as.data.frame(x)))
colnames(nullframe)<-c("x","y","ice")
```

```{r}
allice<-bind_rows(list(data.frame(ice=mxy$ice,PA=1,type="Observed"),data.frame(ice=nullframe$ice,PA=0,type="Random")))

#as density plot
allice$type<-factor(allice$type,levels=c("Random","Observed"))
ggplot(data=allice) + geom_density(aes(x=ice,fill=type),alpha=0.9) + theme_bw() + scale_alpha(range=c(0.5,1)) + scale_fill_manual(values=c("grey90","grey20")) + labs(fill="Data",x="Sea Ice Concentration (%)")

#as cumlative distribution
#ggplot(allice,aes(x=ice,col=type)) + stat_ecdf()

#create PA frame
allice$PA<-(allice$type=="Observed")*1
  
ggplot(allice,aes(x=ice,y=PA)) + geom_smooth(method="glm",method.args=list(family="binomial")) + geom_point() + labs(x="Sea ice concentration",y="Presence")                    
```

#Distance to ice edge

```{r}
#unique days to pull raster
udays<-unique(strptime(mxy$timestamp,format="%Y-%m-%d"))
mxy$iceday<-strptime(mxy$timestamp,format="%Y-%m-%d")

#for each iceday, get the sea ice concentration
out<-list()

cl<-makeCluster(4,"SOCK")
registerDoSNOW(cl)
out<-foreach(x=1:length(udays),.packages="raster") %dopar% {
  print(x)
  #split into components
  mn<-format(udays[x],format="%m")
  
  #month abbreciation
  mnabb<-tolower(month.abb[as.numeric(mn)])
  
  day<-format(udays[x],format="%d")
  year<-format(udays[x],format="%Y")
  
  fl_final<-paste('asi-AMSR2-s3125-',year,mn,day,'.grd',sep="")
  loadfile<-paste("C:/Users/Ben/Dropbox/Whales/AMSR2",year,mnabb,region,fl_final,sep="/")
  
  #load raster
  #tryCatch({
  r<-raster(loadfile)
  
  #ice
  ice<-r>15
  
  #mask that adds coastline
  ice_edge_coast<-ice
  ice_edge_coast[is.na(ice_edge_coast)]<-1
  
  #compute edge
  ice_edge<-boundaries(ice_edge_coast,classes=T)
  
  #extract pts
  upts<-mxy[mxy$iceday %in% udays[x],c("stereo.x","stereo.y")]
  
  #extract distance to each points
  daypoints<-list()
  for(j in 1:nrow(upts)){
    distk<-distanceFromPoints(xy=upts[j,],ice_edge)  
    
    #mask open water
    ice_edge[ice_edge==0]<-NA
    
    #Distance on the ice edge
    distk<-mask(distk,ice_edge)
    
    #Closest cell in km
    daypoints[[j]]<-min(distk[],na.rm=T)/1000
  }
  
  return(data.frame(upts,dist_ice=unlist(daypoints)))
  }
  #, error=function(e){
    #print("No file")
  #  return(NA)
  #})
#}

stopCluster(cl)
#distance to nearest cell
distice_frame<-bind_rows(out)
mxy<-merge(mxy,distice_frame,by=c("stereo.x","stereo.y"))

#If the sea ice conc is > 15%, the ice distance is negative, this isn't true for polynynas need to think about this.
ggplot(mxy,aes(x=dist_ice,y=ice)) + geom_point() + labs(x="Distance to ice edge",y="Sea Ice Concentration")

mxy[mxy$ice>15 & !is.na(mxy$ice),"dist_ice"] <- -mxy[mxy$ice>15 & !is.na(mxy$ice),"dist_ice"]
```

```{r}
#plot distance, with cell size labeled
ggplot(data=mxy,aes(x=dist_ice)) + geom_histogram() + labs(x="Distance to 15% ice edge (km)") + theme_bw() 

ggplot()+geom_point(data=mxy,aes(x=x, y=y,col=dist_ice),size=.5)  + facet_wrap(~Month,nrow = 2)  + borders() + coord_cartesian(ylim = c(-70,-62),xlim=c(-55,-73)) + theme_bw() + mytheme + scale_color_continuous(low="blue",high="red") 
```

### Zoom in 

```{r}
#plot distance, with cell size labeled
ggplot(data=mxy[mxy$dist_ice<100,],aes(x=dist_ice)) + geom_histogram() + labs(x="Distance to ice edge") + theme_bw() 

#By month
ggplot(data=mxy[mxy$dist_ice<100,],aes(x=dist_ice,fill=Month)) + geom_density(alpha=0.5) + labs(x="Distance to ice edge") + theme_bw()

ggplot()+geom_point(data=mxy[mxy$dist_ice<100,],aes(x=x, y=y,col=dist_ice),size=.5)  + facet_wrap(~Month,nrow = 2)  + borders() + coord_cartesian(ylim = c(-70,-62),xlim=c(-55,-73)) + theme_bw() + mytheme + scale_color_continuous(low="blue",high="red") 
```

# Temperal window

Sample the cells in a sliding 10 day window

```{r}
#list all rasters available 
#for each iceday, get the sea ice concentration
allgrd<-list.files("C:/Users/Ben/Dropbox/Whales/AMSR2",recursive = T,full.names = T,pattern=".grd")

out<-list()
for (x in 1:length(allgrd)){
  
  r<-raster(allgrd[x])
  
  #parse date
  dater<-str_match(r@file@name,"s3125-(\\w+)")[,2]
  
  yr<-substring(dater,1,4)
  mn<-substring(dater,5,6)
  dy<-substring(dater,7,8)
  
  #construct date
  set_date<-as.POSIXct(paste(yr,mn,dy,sep="-"),format="%Y-%m-%d",tz="GMT")
  
  gpts<-mxy$iceday[mxy$iceday >= set_date - (10*daysec) & mxy$iceday <= set_date + (10*daysec)]
  
  if(length(gpts)==0){
    next
  }
  #animal location and dates
  upts<-mxy[mxy$iceday %in% gpts,c("stereo.x","stereo.y")]
  upts_date<-mxy[mxy$iceday %in% gpts,c("iceday")]
  upts_ice<-mxy[mxy$iceday %in% gpts,c("ice")]
  upts_animal<-mxy[mxy$iceday %in% gpts,c("Animal")]

  out[[x]]<-data.frame(Animal=upts_animal,upts,obs_day=upts_date,raster_day=set_date,obs_ice=upts_ice,ice=raster::extract(x=r,y=SpatialPoints(cbind(upts))))
  }

ice_frame<-bind_rows(out)

#difference in the days
ice_frame$daydiff<-difftime(ice_frame$obs_day,ice_frame$raster_day,units="days")

#differnce in ice
ice_frame$icediff<-ice_frame$obs_ice-ice_frame$ice

#month
ice_frame$Month<-months(ice_frame$obs_day)
ice_frame$Month<-factor(ice_frame$Month,levels=month.name)


ggplot(ice_frame,aes(x=as.numeric(daydiff),y=ice)) + geom_smooth() + labs(y="Ice concentration",x="Days since observation") + facet_wrap(~Month,scales="free")

ggplot(ice_frame,aes(x=as.numeric(daydiff),y=icediff)) + geom_boxplot(aes(group=daydiff)) + labs(y="Change in ice concentration",x="Days since observation")

ggplot(ice_frame,aes(x=as.numeric(daydiff),y=icediff)) + geom_smooth() + labs(y="Change in ice concentration",x="Days since observation") + facet_wrap(~Month,scales="free")

ggplot(ice_frame,aes(x=as.numeric(daydiff),y=icediff)) + geom_smooth(aes(group=Animal)) + labs(y="Change in ice concentration",x="Days since observation") + facet_wrap(~Month,scales="free") + theme_bw()

```

```{r}
save.image("Data/IceData.RData")
```

